MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-06-03 MAESTROeX_I/algorithm.tex
Latest plots and updates

We shall refer to local atmospheric flows in two and three dimensions as problems in ``planar'' geometry, and full-star flows
in three dimensions as problems in ``spherical'' geometry. The solution in both cases consists of the Cartesian grid solution
and the one-dimensional base state solution.
Figure \ref{Fig:BaseGrid} illustrates the relationship between the base state and the Cartesian grid state for both planar and spherical geometries in the presence of spatially adaptive grids. \MarginPar{did this fig come from one of the earlier papers? if so, we will need to acknowledge that paper and make sure ApJ knows.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{./figs/base_spherical}
\caption{\label{Fig:BaseGrid}  
(Left) For multi-level problems in planar geometry, we force a direct alignment
between the radial array cell centers and the Cartesian grid cell centers by 
allowing the radial base state spacing to change with space and time.
(Right) For multi-level problems in spherical geometry, since there is no direct alignment
between the radial array cell centers and the Cartesian grid cell centers, we choose to fix
the radial base state spacing across levels.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the key numerical modules is the ``lateral average'', which computes the average over a 
layer of a Cartesian grid variable onto a one-dimensional radial array.
In planar geometries, this is a straightforward arithmetic average of cells at
a particular height since the radial cell centers are in alignment
with the Cartesian grid cell centers.
However for spherical problems, the procedure is much more complicated.
In Section 4 of Paper V, we describe how there is a finite, easily computable set of radii that any Cartesian cell-center can map to.  
Specifically, for every Cartesian cell, there exists an integer $m$ such that the radius to the center of the star is given by $\hat{r}_m=\Delta x\sqrt{0.75+2m}$.
We average by binning all the cells associated with each radii and taking the average.
Then we interpolate this data onto a one-dimensional radial array.  
Previously, MAESTRO only allowed for radial arrays with a constant $\Delta r$ (typically equal to $\Delta x/5$).
Here we present a new option to retain an irregularly-spaced radial array to eliminate mapping errors back onto the Cartesian grid state.
Consider a spherical star in hydrostatic equlibrium at rest.  In the absense of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced array, the effects due to interpolation error are completely eliminated.
\MarginPar{Doreen you had a great figure of this mapping in one of your talks}

@@

First, we compute the time-centered edge velocities, $\Ub^{\nph,\pred}$
using a Godunov approach \citep{XRB_III}. Then, we define
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
We update the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{equation}
\Ub^{n+1,\dagger}
= \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right]
 - \dt \left[ \frac{1}{\rho^\nph} \nabla \pi^\nmh + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],
\end{equation}
Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
We use an approximate projection to project $\Ub^{n+1,\dagger}$ onto the space of velocities that satisfy the constraint to obtain $\Ub^{n+1}$ using a ``nodal'' projection.
This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.
Furthermore, as part of the nodal projection, we also define a nodal new-time perturbational pressure, $\pi^\nph$.
Refer to Appendix \ref{Sec:Projection} for more details.

>>>>

We shall refer to local atmospheric flows in two and three dimensions as problems in ``planar'' geometry, and full-star flows
in three dimensions as problems in ``spherical'' geometry. The solution in both cases consists of the Cartesian grid solution
and the one-dimensional base state solution.
Figure \ref{Fig:BaseGrid} illustrates the relationship between the base state and the Cartesian grid state for both planar and spherical geometries in the presence of spatially adaptive grids. \MarginPar{did this fig come from one of the earlier papers? if so, we will need to acknowledge that paper and make sure ApJ knows.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_grid} \hspace{0.5in}
\includegraphics[height=2.0in]{./figs/base_spherical}
\caption{\label{Fig:BaseGrid}  
(Left) For multi-level problems in planar geometry, we force a direct alignment
between the radial array cell centers and the Cartesian grid cell centers by 
allowing the radial base state spacing to change with space and time.
(Right) For multi-level problems in spherical geometry, since there is no direct alignment
between the radial array cell centers and the Cartesian grid cell centers, we choose to fix
the radial base state spacing across levels.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One of the key numerical modules is the ``lateral average'', which computes the average over a 
layer of a Cartesian grid variable onto a one-dimensional radial array.
In planar geometries, this is a straightforward arithmetic average of cells at
a particular height since the radial cell centers are in alignment
with the Cartesian grid cell centers.
However for spherical problems, the procedure is much more complicated.
In Section 4 of Paper V, we describe how there is a finite, easily computable set of radii that any Cartesian cell-center can map to.  
Specifically, for every Cartesian cell, there exists an integer $m$ such that the radius to the center of the star is given by $\hat{r}_m=\Delta x\sqrt{0.75+2m}$.
We average by binning all the cells associated with each radii and taking the average.
Then we interpolate this data onto a one-dimensional radial array.

Previously, MAESTRO only allowed for radial arrays with a constant $\Delta r$ (typically equal to $\Delta x/5$).
Here we present a new option to retain an irregularly-spaced radial array to eliminate mapping errors back onto the Cartesian grid state.
Consider a spherical star in hydrostatic equlibrium at rest.  In the absense of reactions, the star should remain at rest.
The buoyancy forcing term in the momentum equation contains $\rho-\rho_0$.  With the original scheme, interpolation errors in computing $\rho_0$ by averaging would cause artificial acceleration in the velocity field due to the interpolation error from the Cartesian grid to and from the radial base state.  By retaining the radial base state as an irregularly spaced array, the effects due to interpolation error are completely eliminated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tb]
\centering
\includegraphics[height=2.0in]{./figs/base_spherical_new}
\caption{\label{Fig:NewBaseGrid}  
A direct mapping between the radial array cell centers (red squares) and the Cartesian grid cell centers (blue crosses) 
is enforced by computing the average of the grid cell centers that share the same radial distance from 
the center of the domain.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

First, we compute the time-centered edge velocities, $\Ub^{\nph,\pred}$
using a Godunov approach \citep{XRB_III}. Then, we define
\begin{equation}
\rho^\nph = \frac{\rho^n + \rho^{n+1}}{2}, \qquad \rho_0^\nph = \frac{\rho_0^n + \rho_0^{n+1}}{2}.
\end{equation}
We update the velocity field $\Ub^n$ to $\Ub^{n+1,\dagger}$ by discretizing
equation (\ref{eq:momentum}) as
\begin{equation}
\Ub^{n+1,\dagger}
= \Ub^n - \dt \left[\uadvtwo \cdot \nabla \Ub^{\nph,\pred} \right]
 - \dt \left[ \frac{\beta_0^\nph}{\rho^\nph} \nabla \left( \frac{\pi^\nmh}{\beta_0^\nmh} \right) + \frac{\left(\rho^\nph-\rho_0^\nph\right)}{\rho^\nph} g^{\nph} \eb_r \right],
\end{equation}
Again, the $\dagger$ superscript refers
to the fact that the updated velocity does not satisfy the divergence constraint,
\begin{equation}
\nabla \cdot \left(\beta_0^{n+1} \Ub^{n+1} \right) = \beta_0^{n+1} \left[ S^{n+1} - \frac{1}{\gammaonebar^{n+1}p_0^{n+1}}\left(\frac{\partial p_0}{\partial t}\right)^{\nph}\right].\label{eq:div3}
\end{equation}
We use an approximate projection to project $\Ub^{n+1,\dagger}$ onto the space of velocities that satisfy the constraint to obtain $\Ub^{n+1}$ using a ``nodal'' projection.
This projection necessarily differs from the MAC projection used in
{\bf Step 3} and {\bf Step 7} because the velocities in those steps are defined
on edges and $\Ub^{n+1}$ is defined at cell centers, requiring different divergence
and gradient operators.
Furthermore, as part of the nodal projection, we also define a nodal new-time perturbational pressure, $\pi^\nph$.
Refer to Appendix \ref{Sec:Projection} for more details.
